---
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

# The Border Surge?
### Investigating if there's a true crisis at the border

## Intro
I've decided to start doing all my coding in an Rmarkdown doc - that way I can more easily track through the iterations that it takes to get to a final post, chart, etc. This will be a bit less formal and more loosey goosey than the blog (maybe I should start referrring to this as the blog & anything on the site as articles?), so don't expect anything fancy (or necessarily coherent).

This particular piece was inspired by [this WaPo article](https://www.washingtonpost.com/politics/2021/03/23/theres-no-migrant-surge-us-southern-border-heres-data/), which looked to see if there actually *is* a surge at the border or if the surge is just a combination of seasonality and pent up demand while restricted from travel during 2020 due to COVID-19. 

## Digging into the data
```{r libraries}
library(tidyverse)
```

```{r frame setup}
# creating dataframe
f_encounter <- tibble(fiscal_yr = c(rep("2018", 12),
                                    rep("2019", 12),
                                    rep("2020", 12),
                                    rep("2021", 5)),
                      # month 1 = oct (start of FY)
                      month = c(rep(seq(1, 12, 1), 3),
                                seq(1, 5)),
                      # data input manually...
                      encounters = c(34871, 39051, 40519, 35905, 36751, 50347,
                                     51168, 51862, 43180, 40149, 46719, 50568,
                                     60781, 62469, 60794, 58317, 76545, 103731,
                                     109415, 144116, 104311, 81777, 62707, 52546,
                                     45139, 42643, 40565, 36585, 36687, 34460,
                                     17106, 23237, 33049, 40929, 50014, 57674,
                                     71946, 72111, 74018, 78442, 100441))

# quick plot
f_encounter %>%
  ggplot(aes(x = month,
             y = encounters,
             group = fiscal_yr)) + 
  geom_line()


```

Some notes - this is only a small subset of historical data. Not a lot of history available, unless available somewhere else? Check if there is a database elsewhere with this info

```{r model setup}
library(modelr)

# create model
m_crossing <- lm(encounters ~ month, data = f_encounter)
 
# add predictions & plot
f_encounter %>%
  add_predictions(m_crossing) %>%
  ggplot(aes(x = month, y = pred)) +
  geom_line()

```

Not sure that this is all that good for the seasonal shindig, maybe a nonlinear model? Something to look into

```{r model apply}
# creating another frame with dates...
f_temp <- tibble(fy_month = seq(1, 12, 1),
                 cal_month = c(seq(10, 12, 1), seq(1, 9, 1)))

# joining & plotting
f_encounter %>%
  left_join(f_temp, by = c("month" = "fy_month")) %>%
  mutate(date = as.Date(paste(fiscal_yr, 
                              cal_month,
                              1,
                              sep = "-"))) %>%
  add_predictions(m_crossing) %>%
  ggplot(aes(x = date,
             y = pred)) +
  geom_line()

```

This is a bit askew - it's because there's a mix of calendar year & fiscal year that I need to fix...

```{r frame fix}
# creating new column for calendar year alignment
f_encounter$cal_yr <- c(rep(2017, 3),
                        rep(2018, 12),
                        rep(2019, 12),
                        rep(2020, 12),
                        rep(2021, 2))

# joining & plotting again
f_encounter %>%
  left_join(f_temp, by = c("month" = "fy_month")) %>%
  mutate(date = as.Date(paste(cal_yr, 
                              cal_month,
                              1,
                              sep = "-"))) %>%
  add_predictions(m_crossing) %>%
  ggplot(aes(x = date,
             y = pred)) +
  geom_line()
```

Cool - that fixes that issue. Now I'll make a plot that shows the over/underperformance. There's guaranteed to be quite a bit of performance issues, since this is a 1-var linear model... We'll see how it looks & if it captures any of the seasonal differences. 

```{r model performance}
# assigning frame now that it's fixed.
f_encounter <- f_encounter %>%
  left_join(f_temp, by = c("month" = "fy_month")) %>%
  mutate(date = as.Date(paste(cal_yr, 
                              cal_month,
                              1,
                              sep = "-"))) %>%
  add_predictions(m_crossing)

# adding over/under performance (residual)
f_encounter %>%
  mutate(res = encounters - pred) %>%
  ggplot(aes(x = date,
             y = res)) + 
  geom_line()
```

Coolio. This gives some useful info - 2019 was a huge year for immigration. Should also double check the encounters overall...

```{r encounters check}
# assigning frame from above
f_encounter <- f_encounter %>%
  mutate(res = encounters - pred)

# plot 
f_encounter %>%
  ggplot(aes(x = date,
             y = encounters)) + 
  geom_line()

```

Hmm... Not sure if there's actually anything useful there. It doesn't look like there is a lot of difference (except with scale) between the two plots... 

Ah okay I founda  new dataset going back to 2012 - may be able to fit a better model. 

```{r new frame setup}
# rebuilding old frame, ground up...
f_encounter <- tibble(fiscal_yr = c(rep("2012", 12),
                                    rep("2013", 12),
                                    rep("2014", 12),
                                    rep("2015", 12),
                                    rep("2016", 12),
                                    rep("2017", 12),
                                    rep("2018", 12),
                                    rep("2019", 12),
                                    rep("2020", 12),
                                    rep("2021", 5)),
                      month = c(rep(seq(1, 12, 1), 9),
                                seq(1, 5)),
                      encounters = c(31323, 28601, 24360, 30758, 36990, 48520,
                                     46660, 42995, 36794, 32801, 33686, 32375,
                                     34836, 33153, 29075, 32481, 40632, 54009, 
                                     54761, 50481, 40785, 39993, 41110, 38182,
                                     41828, 38685, 36695, 35181, 42399, 57405, 
                                     59119, 68804, 66541, 48819, 39758, 34003,
                                     35895, 33023, 34238, 30178, 32550, 39159,
                                     38296, 40681, 38616, 38610, 42414, 41165,
                                     45507, 45752, 48737, 33654, 38309, 46117, 
                                     48502, 55442, 45772, 46966, 51961, 56535,
                                     66708, 63361, 58412, 42463, 23555, 16588, 
                                     15766, 19940, 21657, 25019, 30567, 31155,
                                     34871, 39051, 40519, 35905, 36751, 50347,
                                     51168, 51862, 43180, 40149, 46719, 50568,
                                     60781, 62469, 60794, 58317, 76545, 103731,
                                     109415, 144116, 104311, 81777, 62707, 52546,
                                     45139, 42643, 40565, 36585, 36687, 34460,
                                     17106, 23237, 33049, 40929, 50014, 57674,
                                     71946, 72111, 74018, 78442, 100441))

# adding in cal year and cal month
f_encounter$cal_yr <- c(rep(2011, 3),
                        rep(2012, 12),
                        rep(2013, 12),
                        rep(2014, 12),
                        rep(2015, 12),
                        rep(2016, 12),
                        rep(2017, 12),
                        rep(2018, 12),
                        rep(2019, 12),
                        rep(2020, 12),
                        rep(2021, 2))

f_encounter <- f_encounter %>%
  left_join(f_temp, by = c("month" = "fy_month")) %>%
  mutate(date = as.Date(paste(cal_yr, cal_month, 1, sep = "-")))

# quick plot
f_encounter %>%
  ggplot(aes(x = date,
             y = encounters)) +
  geom_line()

```

Okay that looks a lot better - can clearly see at least some sembelance of a repeating pattern with spikes near the beginning of years. Wonder why that is. 

Back to trying to fit a linear model (which, again, isn't necessarily a good representation of how this works). First need to do a monthly plot.

```{r new monthly plot}
# create plot
f_encounter %>%
  ggplot(aes(x = cal_month,
             y = encounters,
             group = cal_yr)) +
  geom_line()
```

Sweet, that's exactly what I wanted. I'll note that there are two little "sticks", one on the left and one on the right. That's because the data is reported for the fiscal year, which starts in October 2011 (explains the stick on the right) and I only have data up to Feb 2021 (explains the stick on the left). 

Now I'll try to fit another linear model - hopefully, with 5 extra years of data, it fits a bit better. 2019 & 2020 look to be off years, so I should have known that trying to fit a model to just that subset would have weird/uninformative results. That being said, it's really only 60 extra data points, so I'm a bit apprehensive as to whether or not it'll be worth anything...

```{r new model}
# fitting a new model based on the updated frame
m_crossing <- lm(encounters ~ cal_month, data = f_encounter)

# adding preds & plotting
f_encounter %>%
  add_predictions(m_crossing) %>%
  ggplot(aes(x = date,
             y = pred)) +
  geom_line()

```

Alright - now we've got this seasonal model (again, that's not the greatest) going back ~9 years. Let's see what happens when we plot the residuals.

```{r resids}
# plotting residuals
f_encounter %>%
  add_residuals(m_crossing) %>%
  ggplot(aes(x = date,
             y = resid)) + 
  geom_line()

```

Hmm.... That's still pretty ugly. Let me try something new.

I've been treating the month as a continuous variable, which is why the model spits out this super sharp drop from December to January. I really should treat it as a *categorical* variable. Let's see if that helps:

```{r new model again}
# new categorical model
m_crossing <- lm(encounters ~ factor(cal_month), data = f_encounter)

# new model prediction plot
f_encounter %>%
  add_predictions(m_crossing) %>%
  ggplot(aes(x = date,
             y = pred)) +
  geom_line()

```

Okay, this is much more useful - this captures that seasonal change (and does so better than a one off line that I got from the original linear model). I'll still note that this isn't necessarily a *great* model, but I think this gets us a lot closer to the finish line. 

Now let's plot the residuals.

```{r plotting resids}
# residual plot
f_encounter %>%
  add_residuals(m_crossing) %>%
  ggplot(aes(x = date, 
             y = resid)) + 
  geom_line()
```

Okay, disappointment. The residuals still don't show any long term trends... I wonder what it would look like if we fit a new model based off data from *before* Trump's inauguration and see if that has any better explanatory power?

```{r filtered frame}
f_encounter %>%
  filter(date <= as.Date("2017-01-01")) %>%
  ggplot(aes(x = cal_month,
             y = encounters,
             group = cal_yr)) +
  geom_line()
```

It's a bit more structured, so now let's see if a new model based on this helps...

```{r yet another model}
# recreating & saving last frame
f_encounter_pre <- f_encounter %>%
  filter(date <= as.Date("2017-01-01"))

# fitting a new model
m_crossing_pre <- lm(encounters ~ factor(cal_month), data = f_encounter_pre)

# plotting new model's predictions
f_encounter_pre %>%
  add_predictions(m_crossing_pre) %>%
  ggplot(aes(x = date,
             y = pred)) +
  geom_line()
```

Okay, I think this is a bit more representative. The *huge* spike in 2019 was giving a giant mid-year boost to the model. I'm still not sure if cutting off the data like this is good practice, but let's see what the results are.

First I'll check this model against the 2011-2016 dataset.

```{r new model check}
# model residual plot
f_encounter_pre %>%
  add_residuals(m_crossing_pre) %>%
  ggplot(aes(x = date,
             y = resid)) +
  geom_line()
```

I'm gonna guess that the residual plot of this model against the full dataset is going to be disappointing...

```{r another residual plot}
# adding residual & time plot in same chunk just to compare
f_encounter %>%
  add_residuals(m_crossing_pre) %>%
  ggplot(aes(x = date,
             y = resid)) +
  geom_line()

f_encounter %>%
  ggplot(aes(x = date,
             y = encounters)) +
  geom_line()
```

Wow. I'm really not seeing the seasonality described in the WaPo article. It's hard to tell if it's because the model is too simple (just month as a predictor) or if there really isn't a seasonality effect. On the one hand, there are quite a few years where there's an up-tick early in the year, but there are also years where there's a *drop* at the beginning of the year. There are definitely a lot of other factors that don't get captured here, like policy decisions, country of origin (and factors from *within* that country that would affect someones decision to migrate), etc. 

I think that there are two key things that can be taken away from this:

1. Seasonality *on it's own* is definitely not a strong predictor of apprehensions at the border.
2. Relative to the data from 2012-2017, 2019, 2020, and 2021 (already) have months with extremely high number of border crossings. 

I think that this *lack* of a clear relation between the month & encounters is worth talking about, since it's a refutation of the WaPo article. 

Well, okay, before I do that, I want to check out the 3 month and 6 month averages. I don't necessarily think that it'll add any insight, but on the off chance that I'm wrong...

```{r avg}
# adding zoo for rolling numbers
library(zoo)

f_encounter %>%
  mutate(movavg_3 = rollmean(encounters,
                             k = 3,
                             align = "right",
                             fill = NA),
         movavg_6 = rollmean(encounters,
                             k = 6,
                             align = "right",
                             fill = NA)) %>%
  ggplot() +
  geom_line(aes(x = date,
                y = movavg_3),
            color = "red") +
  geom_line(aes(x = date,
                y = movavg_6),
            color = "blue") +
  labs(title = "3/6 month rolling averages",
       subtitle = "3 month in red, 6 month in blue")

```

Alrighty, now to create *more* models:

```{r more models...}
# adding rollmeans to frame.
f_encounter <- f_encounter %>%
  mutate(movavg_3 = rollmean(encounters,
                             k = 3,
                             align = "right",
                             fill = NA),
         movavg_6 = rollmean(encounters,
                             k = 6,
                             align = "right",
                             fill = NA))

# adding in models
m_crossing_3ma <- lm(movavg_3 ~ factor(cal_month), data = f_encounter)
m_crossing_6ma <- lm(movavg_6 ~ factor(cal_month), data = f_encounter)

# adding in pred plots
f_encounter %>%
  add_predictions(m_crossing_3ma) %>%
  ggplot(aes(x = date,
             y = pred)) +
  geom_line()

f_encounter %>%
  add_predictions(m_crossing_6ma) %>%
  ggplot(aes(x = date,
             y = pred)) +
  geom_line()

```

Interesting. Let's see how the residual plots look. I may run through a quick pre-Trump model (before the outlier 2019) after...

```{r more resids}
# residual plots
f_encounter %>%
  add_residuals(m_crossing_3ma) %>%
  ggplot(aes(x = date,
             y = resid)) +
  geom_line()

f_encounter %>%
  add_residuals(m_crossing_6ma) %>%
  ggplot(aes(x = date,
             y = resid)) +
  geom_line()
```

Yeah, much ado about nothing. Okay, last run through with some new models for *pre* Trump (again, don't think this'll show anything different).

```{r more models broooooo}
# updating pre frame & creating new models
f_encounter_pre <- f_encounter %>%
  filter(date <= as.Date("2017-01-01"))

m_crossing_pre_3ma <- lm(movavg_3 ~ factor(cal_month), data = f_encounter_pre)
m_crossing_pre_6ma <- lm(movavg_6 ~ factor(cal_month), data = f_encounter_pre)

# add preds
f_encounter_pre %>%
  add_predictions(m_crossing_pre_3ma) %>%
  ggplot(aes(x = date,
             y = pred)) +
  geom_line() +
  labs(title = "Pred plot",
       subtitle = "3-mo pred")

f_encounter_pre %>%
  add_predictions(m_crossing_pre_6ma) %>%
  ggplot(aes(x = date,
             y = pred)) +
  geom_line() + 
  labs(title = "Pred plot",
       subtitle = "6-mo pred")

# add resids (pre dataset)
f_encounter_pre %>%
  add_residuals(m_crossing_pre_3ma) %>%
  ggplot(aes(x = date,
             y = resid)) +
  geom_line() +
  labs(title = "Resid plot",
       subtitle = "3-mo resid")

f_encounter_pre %>%
  add_residuals(m_crossing_pre_6ma) %>%
  ggplot(aes(x = date,
             y = resid)) +
  geom_line() +
  labs(title = "Resid plot",
       subtitle = "6-mo resid")

# add resids (full dataset based on pre- model)
f_encounter %>%
  add_residuals(m_crossing_pre_3ma) %>%
  ggplot(aes(x = date,
             y = resid)) +
  geom_line() +
  labs(title = "Resid plot",
       subtitle = "3-mo resid")

f_encounter %>%
  add_residuals(m_crossing_pre_6ma) %>%
  ggplot(aes(x = date,
             y = resid)) +
  geom_line() +
  labs(title = "Resid plot",
       subtitle = "6-mo resid")
```

Beep boop, kind of what I thought. Alright, now just to make the final plots.

## Putting together the final plots

Firstly, I need to create the baseline plot - border encounters over time. 

```{r}
# adding in items for themes/formatting
library(ggtext)
extrafont::loadfonts(device = "win")
devtools::source_url('https://raw.githubusercontent.com/markjrieke/thedatadiary/main/dd_theme_elements/dd_theme_elements.R')

# uptick since the election
p_1 <- f_encounter %>%
  filter(date >= as.Date("2020-11-1")) %>%
  ggplot(aes(x = date, 
             y = encounters)) +
  geom_line(size = 1.1,
            color = dd_blue_dark) +
  labs(title = "Encounters at the border are on the rise",
       subtitle = "Total Southwest Land Border Encounters since November 2020",
       caption = "Data from US Border Patrol (USBP) and Office of Field Operations (OFO) dashboard.",
       x = NULL,
       y = NULL) +
  dd_theme +
  theme(plot.title = element_markdown(family = "Siemens Slab", size = 18),
        plot.subtitle = element_markdown(family = "Siemens Slab", size = 14),
        axis.text = element_markdown(family = "Siemens Slab"),
        plot.caption = element_markdown(family = "Siemens Slab"),
        plot.title.position = "plot") +
  scale_y_continuous(labels = scales::comma_format()) +
  geom_point(aes(x = as.Date("2021-02-01"),
                 y = 100441),
             size = 5,
             stroke = 2,
             color = dd_blue_dark,
             fill = dd_cream,
             shape = 21) +
  annotate(geom = "label",
           label = "100,441 encounters\nin February",
           label.size = 0,
           family = "Siemens Slab",
           fontface = "bold",
           fill = dd_orange,
           alpha = 0.25,
           x = as.Date("2021-1-1"),
           y = 95000) +
  annotate(geom = "curve",
           x = as.Date("2021-1-7"),
           y = 97000,
           xend = as.Date("2021-1-28"),
           yend = 100400,
           curvature = -0.2,
           size = 1.1,
           arrow = arrow(length = unit(3, "mm")))

ggsave("p_1.png",
       width = 9,
       height = 6,
       units = "in",
       dpi = 500)

p_1

p_2 <- f_encounter %>%
  ggplot(aes(x = date,
             y = encounters)) +
  geom_line(size = 1.1,
            color = dd_blue_dark) +
  labs(title = "Seasonal spikes seem to appear early in each year",
       subtitle = "Total Southwest Land Border Encounters since October 2011",
       caption = "Data from US Border Patrol (USBP) and Office of Field Operations (OFO) dashboard",
       x = NULL,
       y = NULL) +
  dd_theme +
  theme(plot.title = element_markdown(family = "Siemens Slab", size = 18),
        plot.subtitle = element_markdown(family = "Siemens Slab", size = 14),
        axis.text = element_markdown(family = "Siemens Slab"),
        plot.caption = element_markdown(family = "Siemens Slab"),
        plot.title.position = "plot") +
  scale_y_continuous(labels = scales::comma_format()) +
  annotate(geom = "label",
           label = "Peak of 144,116 \nencounters in May '19",
           label.size = 0,
           family = "Siemens Slab",
           fontface = "bold",
           fill = dd_orange,
           alpha = 0.25,
           x = as.Date("2017-1-1"),
           y = 110000) +
  annotate(geom = "curve",
           x = as.Date("2017-5-1"),
           y = 117500,
           xend = as.Date("2019-3-1"),
           yend = 144000,
           curvature = -0.2,
           arrow = arrow(length = unit(3, "mm")),
           size = 1.1,
           curvature = -0.2)

ggsave("p_2.png",
       width = 9,
       height = 6,
       units = "in",
       dpi = 500)

p_2

p_3 <- f_encounter %>%
  ggplot(aes(x = cal_month,
             y = encounters,
             group = cal_yr,
             color = as.character(cal_yr),
             alpha = as.character(cal_yr))) +
  geom_line(size = 1.1) +
  scale_color_manual(values = c(rep(dd_gray, 8),
                                dd_red,
                                dd_blue,
                                dd_green)) +
  scale_alpha_manual(values = c(rep(0.4, 8),
                                rep(1, 3))) +
  dd_theme + 
  theme(plot.title = element_markdown(family = "Siemens Slab", size = 18),
        plot.subtitle = element_markdown(family = "Siemens Slab", size = 14),
        axis.text = element_markdown(family = "Siemens Slab"),
        plot.caption = element_markdown(family = "Siemens Slab"),
        plot.title.position = "plot",
        legend.position = "none") +
  labs(title = "Recent breaks from the trend",
       subtitle = "Monthly encounters in <span style = 'color:#D75565;'>**2019,**</span> <span style = 'color:#5565D7;'>**2020,**</span> and <span style = 'color:#65D755;'>**2021**</span> differ from previous years",
       x = NULL,
       y = NULL,
       caption = "Data from US Border Patrol (USBP) and Office of Field Operations (OFO) dashboard") + 
  scale_y_continuous(labels = scales::comma_format()) +
  scale_x_continuous(breaks = seq(1, 12, 2),
                     labels = c("Jan", "Mar", "May", "Jul", "Sep", "Nov"))

ggsave("p_3.png",
       width = 9,
       height = 6,
       units = "in",
       dpi = 500)

p_3
```

As an interjection... I reknit the doc & was scrolling through. It looks like there's both annual over/under performance of the model. I don't necessarily think this will reveal anything new, but I feel like I ought to check a 12 month average. This would still capture seasonality, but should also account for any of the "pent up demand" mentioned in the WaPo article. 

## Hang on - more modeling

Let's do some quick remodeling of 12 month data:

```{r}
# add new col
f_encounter <- f_encounter %>%
  mutate(movavg_12 = rollmean(encounters,
                              k = 12,
                              align = "right",
                              fill = NA))

# add new model
m_crossing_12ma <- lm(movavg_12 ~ factor(cal_month), data = f_encounter)

# plot check preds
f_encounter %>%
  add_predictions(m_crossing_12ma) %>%
  ggplot(aes(x = date,
             y = pred)) + 
  geom_line()

# plot check resids
f_encounter %>%
  add_residuals(m_crossing_12ma) %>%
  ggplot(aes(x = date,
             y = resid)) + 
  geom_line()

```

Yeah, okay that really just makes all the residual plots smoother, doesn't really add any extra info. I kinda thought that at the outset, but good to double check. Continuing on with the final plots:

## Final Plots (Continued)

```{r}
p_4 <- f_encounter %>%
  add_residuals(m_crossing) %>%
  ggplot(aes(x = date,
             y = resid)) +
  geom_line(size = 1.1,
            color = dd_purple) +
  dd_theme +
  labs(title = "Seasonality offers little insight",
       subtitle = "Residual error of expected encounters using month as predictor",
       x = NULL,
       y = NULL,
       caption = "Data from US Border Patrol (USBP) and Office of Field Operations (OFO) dashboard") +
  theme(plot.title = element_markdown(family = "Siemens Slab", size = 18),
        plot.subtitle = element_markdown(family = "Siemens Slab", size = 14),
        axis.text = element_markdown(family = "Siemens Slab"),
        plot.caption = element_markdown(family = "Siemens Slab"),
        plot.title.position = "plot",
        legend.position = "none") +
  scale_y_continuous(labels = scales::comma_format(),
                     breaks = seq(-25000, 75000, 25000))

ggsave("p_4.png",
       width = 9,
       height = 6,
       units = "in",
       dpi = 500)

p_4
  
```

That should do it. I'll note that the way that Rmarkdown renders images is slightly different from the way that ggsave renders images, so things might look a bit odd in this document. Since I'm adding the .pngs to the post, and this is really just a walkthrough/stream of consciousness, I'm not going to worry about it too much. 

